Cluster Difference Imaging Photometric Survey (CDIPS) II. TOI$\,$837: A Young Validated Saturn-Sized Planet in IC~2602

### 2020-07-21 paper/ms.tex
Update manuscript
Co-authored-by: Andrew Mann <awmann@unc.edu>

\begin{abstract}
  We report the discovery of TOI 837.01 and its validation as a
  transiting planet.  We characterize the system using data from the
  NASA TESS mission, the ESA Gaia mission, ground-based photometry
  from El Sauce and ASTEP400, and spectroscopy from the CHIRON, FEROS,
  and Veloce spectrographs.  We find that TOI$\,$837 is a $T=9.9\,{\rm
  mag}$ G0/F9 dwarf in the southern open cluster IC 2602.  The star
  and planet are therefore $35^{+11}_{-5}$ million years old.
  Combining the transit photometry with a prior on the stellar
  parameters derived from the cluster color-magnitude diagram, we find
  that the planet is warm ($P = 8.3\,{\rm d}$) and slightly smaller
  than Jupiter ($R_{\rm p} = 0.80^{+0.09}_{-0.03} \,R_{\rm Jup}$).
  The transits are either grazing or nearly grazing.  From radial
  velocity monitoring, we limit the minimum mass to less than 1.20
  $M_{\rm Jup}$ (3$\sigma$).  Grazing transits are a cause for
  concern, as they are usually indicative of astrophysical false
  positive scenarios.  Our follow-up data show that such scenarios are
  highly unlikely.  Our combined multi-color photometry,
  high-resolution imaging, and radial velocities rule out hierarchical
  eclipsing binary scenarios.  Background eclipsing binary scenarios,
  though limited by speckle imaging, formally remain a 0.2\%
  possibility.  TOI 837.01 is therefore a validated adolescent
  exoplanet.  Further observations of its stellar obliquity,
  atmosphere, and mass should confirm its planetary nature beyond
  doubt, and could improve our understanding of how the physical and
  orbital properties of exoplanets change in time.
\end{abstract}

@@

Finding and studying systems undergoing these evolutionary changes is
a major goal in contemporary exoplanet research.  To identify stars
younger than say 1$\,$Gyr, a number of direct and indirect methods are
available \citep{soderblom_ages_2010}.  The traditional approach is to
isochronally age-date coeval groups of stars, hereafter ``clusters''
\citep[{\it
e.g.},][]{lada_embedded_2003,zuckerman_young_2004,krumholz_star_2019}.
Young field stars can also be identified isochronally, provided that
they are sufficiently massive \citep{berger_gaia-kepler_2020}.  Other
age indicators include stellar rotation periods, the abundance of
photospheric lithium, and chromospheric diagnostics such as calcium
emission and broadband UV emission.  Studies by for instance
\citet{david_discovery_2018} and G{.}~Zhou et al{.}~(2020, submitted)
have combined these methods to age-date individual field stars hosting
transiting planets.  Many of these latter methods were summarized by
\citet{mamajek_improved_2008}, and have since been calibrated by {\it
e.g.},
\citet{irwin_rotational_2009,barnes_color-period_2015,meibom_spin-down_2015,angus_calibrating_2015}
and \citet{curtis_tess_2019} for stellar rotation,
\citet{zerjal_chromospherically_2017} for chromospheric activity, and
{\it e.g.}, \citet{elliott_search_2016,berger_identifying_2018} and
\citet{zerjal_galah_2019} for lithium abundances.

@@

\paragraph{Acquisition and reduction}
We observed four transits with the 0.36$\,$m at Observatario El Sauce,
located in the R\'io Hurtado Valley in Chile, and operated by
co-author P{.}~Evans.  The observations were obtained in Cousins-R
band on the nights of 1 April 2020 and 26 April 2020, Cousins-I band
on the night of 21 May 2020, and Johnson-B band on the night of 14
June 2020.  The final 14 June transit began shortly after twilight.

@@

If the transit were box-shaped, this argument could be extended to
even more restrictive depths \citep[{\it
e.g.},][]{seager_unique_2003,vanderburg_hr858_2019,rizzuto_tess_2020}.
Since the transits of \tn\ are grazing, the second and third
contact points do not occur, and the shape of the transit is not
restrictive.

@@

\subsubsection{Not SB2}
We determined our detection limits to blended spectroscopic companions
using the CHIRON spectra (see Section~\ref{subsec:chiron}).  For a
slowly rotating stellar companion well-separated in velocity, this
would have revealed companions with flux fractions $F_2/F_1 \gtrsim
3\%$.  For a companion with rotational broadening of $15\,{\rm
km\,s}^{-1}$, roughly equivalent to that of \tn, we were able to
exclude companions with flux fractions exceeding $\approx$8\%.  For
plotting purposes, in Figure~\ref{fig:fpscenario} we have assumed the
latter flux-fraction limit of 8\% ($\Delta {\rm mag} \approx 2.7$).
For bound companions this is justified through reasoning that coeval
binary companions to \tn\ would have both longer rotation periods and
smaller stellar radii. The equatorial velocities of such companions
would therefore be expected to be smaller than that of \tn.

@@

Archival SERC-J and AAO-SES plates are available for the \tn\
field\footnote{\url{https://archive.stsci.edu/cgi-bin/dss_form}}.
These plates were acquired in 1982 and 1992, respectively.  For high
proper motion stars archival imagery can be used to detect slowly
moving background stars that might be an astrophysical false-positive
source \citep[{\it
e.g.},][]{bakos_stellar_2006,huang_pimen_2018,vanderburg_hr858_2019}.
However \tn\ has only moved $\approx0.7''$ between 1982 and present,
in comparison to the $\approx2.0''$ FWHM of the target on the plates.
We therefore cannot resolve it with great confidence from background
sources not already resolved through more modern imaging.

@@

Figure~\ref{fig:hr} also compares the data to the MIST isochrones
\citep{choi_mesa_2016}.  We used the
web-interace\footnote{\url{http://waps.cfa.harvard.edu/MIST/interp_isos.html},
\texttt{2020-07-08}} to interpolate isochrones at 10, 20, 30, and 40
million years. We assumed solar metallicity, and a fixed extinction
value of $A_V = 0.217$ \citep{randich_gaiaeso_2018}.  The 30 and 40
Myr models align well with the data for stars with masses ranging from
roughly $0.7$--$7\,M_\odot$.  The PMS K and M dwarf models are bluer
than observed in the Gaia photometry.  This discrepancy was noted and
discussed at length by \citet{choi_mesa_2016}.  The most plausible
explanation seems to be that strong magnetic fields in low-mass
pre-main-sequence stars inhibit convection and produce a high filling
factor of starspots \citep[{\it
e.g.},][]{stauffer_why_2003,feiden_magnetic_2013}. To conserve flux,
the stellar radius expands, and the star becomes cooler than a
1-dimensional model would predict.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f10.png}
	\end{center}
	\vspace{-0.5cm}
	\caption{ 
    {\bf Posterior probabilities of impact parameter, planet-to-star
    size ratio, and stellar density.} The  planet-to-star size ratio
    corresponds to a planet size between $0.68\,R_{\rm Jup}$ and
    $0.97\,R_{\rm Jup}$ (3$^{\rm rd}$--97$^{\rm th}$ percentile).
    This plot was made using \texttt{corner} \citep{corner_2016}.
		\label{fig:subsetcorner}
	}
\end{figure}

@@

A.J.\ and R.B.\ acknowledge support from project IC120009 ``Millennium
Institute of Astrophysics (MAS)'' of the Millenium Science Initiative,
Chilean Ministry of Economy. A.J.\ acknowledges additional support
from FONDECYT project 1171208.

>>>>

\begin{abstract}
  We report the discovery of TOI 837.01 and its validation as a
  transiting planet.  We characterize the system using data from the
  NASA TESS mission, the ESA Gaia mission, ground-based photometry
  from El Sauce and ASTEP400, and spectroscopy from the CHIRON, FEROS,
  and Veloce spectrographs.  We find that TOI$\,$837 is a $T=9.9\,{\rm
  mag}$ G0/F9 dwarf in the southern open cluster IC 2602.  The star
  and planet are therefore $35^{+11}_{-5}$ million years old.
  Combining the transit photometry with a prior on the stellar
  parameters derived from the cluster color-magnitude diagram, we find
  that the planet orbits close to its star ($P = 8.3\,{\rm d}$) and is
  slightly smaller than Jupiter ($R_{\rm p} = 0.80^{+0.09}_{-0.03}
  \,R_{\rm Jup}$).  The transits are either grazing or nearly grazing.
  From radial velocity monitoring, we limit the minimum mass to less
  than 1.20 $M_{\rm Jup}$ (3$\sigma$).  Grazing transits are a cause
  for concern, as they are usually indicative of astrophysical false
  positive scenarios.  Our follow-up data show that such scenarios are
  highly unlikely.  Our combined multi-color photometry,
  high-resolution imaging, and radial velocities rule out hierarchical
  eclipsing binary scenarios.  Background eclipsing binary scenarios,
  though limited by speckle imaging, formally remain a 0.2\%
  possibility.  TOI 837.01 is therefore a validated adolescent
  exoplanet.  Further observations of its stellar obliquity,
  atmosphere, and mass should confirm its planetary nature beyond
  doubt, and could improve our understanding of how the physical and
  orbital properties of exoplanets change in time.
\end{abstract}

@@

Finding and studying systems undergoing these evolutionary changes is
a major goal in contemporary exoplanet research.  To identify stars
younger than say 1$\,$Gyr, a number of direct and indirect methods are
available \citep{soderblom_ages_2010}.  The traditional approach is to
isochronally age-date coeval groups of stars, hereafter ``clusters''
\citep[{\it
e.g.},][]{lada_embedded_2003,zuckerman_young_2004,krumholz_star_2019}.
Young field stars can also be identified isochronally, provided that
they are sufficiently massive \citep{berger_gaia-kepler_2020}.  Other
age indicators include stellar rotation periods, the abundance of
photospheric lithium, and chromospheric diagnostics such as calcium
emission and broadband UV emission.  Studies by for instance
\citet{david_discovery_2018} and G{.}~Zhou et al{.}~(2020, submitted)
have combined these methods to age-date individual field stars hosting
transiting planets.  Many of these latter methods were summarized by
\citet{mamajek_improved_2008}, and have since been calibrated by {\it
e.g.},
\citet{irwin_rotational_2009,barnes_color-period_2015,meibom_spin-down_2015,angus_calibrating_2015}
and \citet{curtis_tess_2019} for stellar rotation,
\citet{zerjal_chromospherically_2017} for chromospheric activity, and
{\it e.g.}, \citet{berger_identifying_2018} and
\citet{zerjal_galah_2019} for lithium abundances.

@@

\paragraph{Acquisition and reduction}
We observed four transits with the 0.36$\,$m at Observatorio El Sauce,
located in the R\'io Hurtado Valley in Chile, and operated by
co-author P{.}~Evans.  The observations were obtained in Cousins-R
band on the nights of 1 April 2020 and 26 April 2020, Cousins-I band
on the night of 21 May 2020, and Johnson-B band on the night of 14
June 2020.  The final 14 June transit began shortly after twilight.

@@

If the transit were box-shaped, this argument could be extended to
even more restrictive depths \citep[{\it
e.g.},][]{seager_unique_2003,vanderburg_hr858_2019,rizzuto_tess_2020}.
Since the transits of \tn\ could be grazing, the second and third
contact points do not necessarily occur, and the shape of the transit
is not particularly restrictive.

@@

\subsubsection{Not SB2}
We determined our detection limits to blended spectroscopic companions
using the stacked CHIRON spectra (see Section~\ref{subsec:chiron}).  For a
slowly rotating stellar companion well-separated in velocity, this
would have revealed companions with flux fractions $F_2/F_1 \gtrsim
3\%$.  For a companion with rotational broadening of $15\,{\rm
km\,s}^{-1}$, roughly equivalent to that of \tn, we were able to
exclude companions with flux fractions exceeding $\approx$8\%.  For
plotting purposes, in Figure~\ref{fig:fpscenario} we have assumed the
latter flux-fraction limit of 8\% ($\Delta {\rm mag} \approx 2.7$).
For bound companions this is justified through reasoning that coeval
binary companions to \tn\ would have both longer rotation periods and
smaller stellar radii. The equatorial velocities of such companions
would therefore be expected to be smaller than that of \tn.

@@

Archival SERC-J and AAO-SES plates are available for the \tn\
field\footnote{\url{https://archive.stsci.edu/cgi-bin/dss_form}}.
These plates were acquired in 1982 and 1992, respectively.  For high
proper motion stars archival imagery can be used to detect slowly
moving background stars that might be an astrophysical false-positive
source \citep[{\it
e.g.},][]{bakos_stellar_2006,huang_pimen_2018,vanderburg_hr858_2019}.
However \tn\ has only moved $\approx0.7''$ between 1982 and present,
in comparison to the $\approx2.0''$ FWHM of the target on the plates.
We therefore cannot resolve it from background sources not already
resolved through more modern imaging.

@@

Figure~\ref{fig:hr} also compares the data to the MIST isochrones
\citep{choi_mesa_2016}.  We used the
web-interace\footnote{\url{http://waps.cfa.harvard.edu/MIST/interp_isos.html},
\texttt{2020-07-08}} to interpolate isochrones at 10, 20, 30, and 40
million years. We assumed solar metallicity, and a fixed extinction
value of $A_V = 0.217$ \citep{randich_gaiaeso_2018}.  The 30 and 40
Myr models align well with the data for stars with masses ranging from
roughly $0.7$--$7\,M_\odot$.  The PMS K and M dwarf models are bluer
than observed in the Gaia photometry.  This discrepancy was noted and
discussed at length by \citet{choi_mesa_2016}.  One suggested
explanation was that strong magnetic fields in low-mass
pre-main-sequence stars inhibit convection and produce a high filling
factor of starspots \citep[{\it
e.g.},][]{stauffer_why_2003,feiden_magnetic_2013}.  This explanation
however fails to explain poor isochrone fits in old open clusters
({\it e.g.}, M$\,$67) as well as the field, particularly in blue
bandpasses. A more likely explanation could be that the molecular line
lists for M dwarf atmospheres are incomplete in these wavelength
ranges.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f10.png}
	\end{center}
	\vspace{-0.5cm}
	\caption{ 
    {\bf Posterior probabilities of impact parameter, planet-to-star
    size ratio, and stellar density.} 
    Contours are shown at 1, 2, 3, and 4-$\sigma$ confidence.
    The planet-to-star size ratio
    corresponds to a planet size between $0.68\,R_{\rm Jup}$ and
    $0.97\,R_{\rm Jup}$ (3$^{\rm rd}$--97$^{\rm th}$ percentile).
    This plot was made using \texttt{corner} \citep{corner_2016}.
		\label{fig:subsetcorner}
	}
\end{figure}

@@

A.J.\ and R.B.\ acknowledge support from project IC120009 ``Millennium
Institute of Astrophysics (MAS)'' of the Millenium Science Initiative,
Chilean Ministry of Economy. A.J.\ acknowledges additional support
from FONDECYT project 1171208.  J.I.V\ acknowledges support from
CONICYT-PFCHA/Doctorado Nacional-21191829.
%

C.Z.\ is supported by a Dunlap Fellowship at the Dunlap Institute for
Astronomy \& Astrophysics, funded through an endowment established by
the Dunlap family and the University of Toronto.
%
